# Data preparation

## Datasets

### Countries ranked by universities (Webometrics)

- Dataset: https://www.webometrics.info/en/distribution_by_country


```{r}
library(knitr)
library(tidyr)
library(dplyr)
library(countrycode)
```
```{r}
# df_ranks <- read.csv('data/webometrics.tsv', sep = '\t')
#
# # Set country rank based on row number
# df_ranks$Rank <- 1:nrow(df_ranks)
```

```{r results = TRUE}
# kable(head(df_ranks, 10))
```

### University ranks (Shanghai rankings)

- Name: Academic Ranking of World Universities (ARWU) - Shanghai Rankings from 2005-2018
- Dataset: https://www.kaggle.com/datasets/joebeachcapital/shanghai-world-university-ranking

```{r}
df_uni <- read.csv('data/shanghai-world-university-ranking.csv',
                   sep = ';',
                   na = "")
# Remove unnecessary columns
df_uni <- df_uni %>% select(-c(Geo.Shape))

# Rename columns
df_uni <- df_uni %>% rename("Rank" = "World.rank.integer")
```
```{r results = TRUE}
kable(head(df_uni %>% select(c("World.rank", "Rank", "University", "Country")), 10))
```
```{r include = FALSE}
str(df_uni)
```
### Human Development Index (HDI)

- Website: https://hdr.undp.org/data-center/human-development-index
- Dataset: https://hdr.undp.org/sites/default/files/2021-22_HDR/HDR21-22_Composite_indices_complete_time_series.csv
- Metadata: https://hdr.undp.org/sites/default/files/2021-22_HDR/HDR21-22_Composite_indices_metadata.xlsx

```{r}
df_hdi <- read.csv('data/HDR21-22_Composite_indices_complete_time_series.csv', sep = ',')

# Select only years which are included in the shanghai uni rankings as well
hdi_years <- 2005:2018

# Select indicators from the HDI group only
df_hdi <- df_hdi %>%
         select(c(iso3, country, hdicode, hdi_rank_2021),
                paste0("hdi_", hdi_years),    # Human Development Index (value)
                paste0("le_", hdi_years),     # Life Expectancy at Birth (years)
                paste0("eys_", hdi_years),    # Expected Years of Schooling (years)
                paste0("mys_", hdi_years),    # Mean Years of Schooling (years)
                paste0("gnipc_", hdi_years),  # Gross National Income Per Capita (2017 PPP$)
         )
```

```{r results = TRUE}
kable(head(df_hdi %>% select(c(iso3, country, hdicode, hdi_rank_2021, hdi_2005, hdi_2018, le_2005, le_2018)), 10))
```
## Preparation

```{r}
# Validate the total score.
Total_score <- function(df) {
  score <- df$Alumni*0.1 + df$Award*0.2 + df$Hi.Ci*0.2 + df$N.S*0.2 + df$PUB*0.2 + df$PCP*0.1
  return(score)
}

df_uni$Total_score <- Total_score(df_uni)
```

### Univerity rank aggregation

Calculate points per for each university based on rank.
```{r}
# Bin universities into ranges and assign points.
df_uni <- df_uni %>%
    mutate(Points = case_when(
        Rank >= 1 & Rank < 10 ~ 100,
        Rank >= 10 & Rank < 20 ~ 50,
        Rank >= 20 & Rank < 50 ~ 25,
        Rank >= 50 & Rank < 100 ~ 10,
        Rank >= 100 ~ 5
  ))
```
Aggregate university ranks per country per year
```{r}
df_uni_agg <- df_uni %>%
    group_by(ISO3.CODE, Country, Year) %>%
    summarise(Points = sum(Points), Count = n()) %>%
    na.omit()
```

### HDI dataset

```{r}
df_hdi_long <- pivot_longer(
    df_hdi,
    cols = -c(iso3, country, hdicode, hdi_rank_2021),
    names_to = "feature",
    values_to = "value")
```
```{r}
# Split feature into indicator and year columns
df_hdi_long <- df_hdi_long %>%
    separate(feature, into = c("indicator", "year"), sep = "_")
df_hdi_long$year <- as.integer(df_hdi_long$year)
```

```{r}
# Transform back to wide format with a column per indicator, but country and year as rows
df_hdi <- pivot_wider(
    df_hdi_long, names_from = "indicator", values_from = "value")
```
```{r}
df_hdi <- left_join(
    df_hdi, df_uni_agg, by = c("year" = "Year", "iso3" = "ISO3.CODE"))

df_hdi <- df_hdi %>%
    select(-Country) %>%
    rename(uni_score = Points, uni_count = Count)
```

Add country region and continent.

```{r}
df_hdi <- df_hdi %>%
    mutate(
        region = countrycode(df_hdi$iso3, origin = "iso3c", destination = "region"),
        continent = countrycode(df_hdi$iso3, origin = "iso3c", destination = "continent"),
        .after = 'country'
    )
```

```{r}
write.csv(df_hdi, 'data/hdi.csv')
```

